\documentclass[12pt]{article}

\usepackage{graphicx}
\usepackage{paralist}
\usepackage{listings}
\usepackage{booktabs}

\oddsidemargin 0mm
\evensidemargin 0mm
\textwidth 160mm
\textheight 200mm

\pagestyle {plain}
\pagenumbering{arabic}

\newcounter{stepnum}

\title{Assignment 1 Solution}
\author{Joshua Guinness, guinnesj}
\date{\today}

\begin {document}

\maketitle

In this assignment, a python program was created which automatically reads in data
from various files, then allocates first year engineering students into their
stream in order of free choice, then GPA. The program consists of three modules; 
one to read in the student, departement, and free choice data, one to allocate
the students and perform operations, and one which acts as a test driver, consisting
of test cases.

The rest of this document will outline the methodology behind the testing
and serve as a reflection on the assignment.

\section{Testing of the Original Program}

In order to talk about testing results and the methodology used to design test cases,
it is important to first talk about what was assumed while coding the program.

The reason assumptions had to be taken is because the specification was not formal or
specific enough on how the interface was meant to work. For example, the specification
gave no instructions on program behaviour when the file does not exist. Either
an assumption had to be made that the file will always exists, or the potential
error had to be dealt with in the code, the later of which was done in this program.

The assumptions used while programming are as follows.

\begin{enumerate}
	\item All the names and macid of the students are correct and match those in
	other files such as freeChoice.
	\item All students have a gender that is either `male' or `female' and the 
	gender is spelt correctly.
	\item All students have a GPA that is between 1.0 and 12.0.
	\item All students have three proper engineering streams that they would like
	to go into, spelt correctly.
	\item There is enough capacity in each department for every student with free choice
	to be allocated to their first choice
	\item There is enough capacity in each department for every student to be allocated
	to at least one of their top three choices.
	\item Every student with free choice passed first year engineering (GPA greater than or equal
	to 4.0).
\end{enumerate}

Based on the assumptions outline above, there were two types of test cases designed.

The first type of test cases checked for correctness of the functions. This type
included the second, third, fourth, fifth, sixth, and eighth test cases. These cases
were chosen to test the main functionality of each of the functions in CalcModule
at least once. After these test cases, it can be confirmed that the implementation
of the functions are correct. A description and rationale of each one will be 
breifly discussed. All these test cases passed.

The second test case checked to see whether the sort function sorted the students
in order of GPA correctly. It just compared the output of the function to
a dictionary of sorted students which was known to be correct. 

The third, fourth, fifth, and sixth test cases check to see whether a correct
average was returned for males and females from two different data sets. The test
case returned false when the exact float value was compared with the output so to
solve the issue, a float value in a very small range was compared with the output.

The eighth test case confirmed that the allocate properly placed students into
their engineering streams first by free choice, then in order of GPA.

The second type of test cases checked for correct behaviour of program when weird
inputs, like missing files, or incomplete data was passed to the functions. This 
type included the first, seventh, and ninth test cases. The number of test cases 
of this type required to build a modest amount of confidence in the program depends
on the amount of assumptions. Since quite a few things were assumed, less test
cases of this type were needed. All these test cases passed after modification to the 
the code. The rationale behind them, problems uncovered while testing, and a brief
description of each test case will be discussed.

In the first test case, the functions in ReadAllocationData were tested instead of
those in CalcModule. It tested for program response when all the files were empty.
The functions responded correctly by returnning empty dictionaries or lists.

In the seventh test case, the average function was checked to see behaviour when
tried to get the average of all the females, but there are no females in the data.
Based on this test case failing for the first time, the code had to be modified to
return 'None', as well as correctly deal with the fact that there is no one of that
gender in the data.

Finally, the ninth test case checked to see whether the allocation function worked 
correctly when it was given student data and none of them had free choice. This was
tested to ensure that the function still operated correctly when one of the files
was empty.

All this list of nine test cases covered quite a bit of program functionality and
possible errors, building confidence in the robustness and correctless of it, 
there is still more testing that could have been done to verify these things. This
was what could be done in the timing given.

%Description of approach to testing.  Rationale for test case selection.  Summary
%of results.  Any problems uncovered through testing.

%Under testing you should list any assumptions you needed to make about the
%program's inputs or expected behaviour.

\section{Results of Testing Partner's Code}

A table outlining the summary of results after running the test script on the
partners code is seen below.

\begin{table}[h]
	\centering
	\begin{tabular}{lr}
		\toprule
		Test Case & Result\\
		\midrule
		1 & Pass\\
		2 & Pass\\
		3 & Pass\\
		4 & Pass\\
		5 & Pass\\
		6 & Pass\\
		7 & Fail - ZeroDivisionError\\
		8 & Pass\\
		9 & Pass\\
		\bottomrule
	\end{tabular}
	\caption{Table outlining the results of testing partner code}
	\label{Table:Table 1}
\end{table}

The results of this table are further discussed in section 3.2

%Summary of results.

\section{Discussion of Test Results}

The following two subsections will discuss and analyze the test results of the code
written by the programmer, and the partners code.

\subsection{Problems with Original Code}

While doing the testing, two problems with the code became evident. The first is that
if there are either no males, or no females, then the average function would divide by 0
and would crash. This was solved by implementing a counter and checking to see if there were
none of that particular gender. The second problem discovered is that in the dictionary of streams
and the students in each stream, each student is supposed to represent a dictionary of a student,
not just the macid. This was again solved by altering the code.

\subsection{Problems with Partner's Code}

As can be seen from the table above all the test cases were passed except for the 
seventh one. The seventh test case check to see how the program behaved when
asked the caluclate the average GPA of all the females, yet there were no females
in the data.

The error that was gotten was a ZeroDivisionError at line 42 in the partners
CalcModule program. This line of code is:

\centerline{\textit{aveFemale = sumFemale / numFemale}}

This line of code divides the total sum by the number of females. However, if there
are zero females, the program tries to divide by zero, throwing an error, leading
the this particular test case to fail.

\section{Critique of Design Specification}

The design specification was good in some ways, but poor in others. 

Since it had been the first time programming in python in a while, having restrictions 
on the types of data structures to use, as well as a laid out module interface made it 
simpler to begin programming. Also, since the specification was in natural language, 
it made it easy to understand, then implement the basic idea of what the modules were 
supposed to do.

The specification was poor in that because it was given in natural language, the
programmer had to make a lot of assumptions about things like operations on the data
and how to deal with errors. The specification should have been made more specific
and explicit in how exactly what the interace was so less assumptions would have to be made.

A proposed change for the design is to give the specification both in natural language and
in a formal way. This was the basic idea of the program can be understood easily, and
the specifics can also be understood how to be dealt with. Specifying the implentation,
especially for a first assignment is good and should be kept that way.

%\newpage

\section{Answers to Questions}

\begin{enumerate}[(a)]

\item The average function can be made more general by instead calculating the average
GPA of students that possess a specific characteristic in common. Right now, the average
function just gets the average of either the males or females, but a more general function
would be able to return the average of all students with the same first stream in common
for example.

Along the same line of thinking, a more general sort function would be able to sort students
by any specified characteristic like last name, macid, or first stream choice.

\item In this context aliasing means you can have multiple names for the same dictionary.
In general, aliasing could be a concern with dictionaries because the dictionary could
be accidently modified or changed due to not knowing the same data structure is pointed
to by two different names. This can be guarded against in two ways. The first is by using
a non-mutable data stucture, like a tuple. This way it cannot be modified, only read from. The
second is by creating copies of the dictionary and only modifiying the copies as needed.

\item Even though the assignment did not require testing of ReadAllocationData, the 
first test case tested how the function within that module respond to empty files. 
It was tested because it was thought that it would be important to ensure that the
functions responsible for getting the data work properly and perform as intended

However some more test cases that could be done to build confidence in ReadAllocationData
are:

\begin{itemize}
\item No filenames are passed to function.
\item Filenames that are passed to function are incorrect.
\item Files that are passed to the function are empty.
\item Files that contain the correct data, but in a format that is not expected
by the function, e.g. tabs instead of newline characters, or missing the students 
macid.
\item Files that are in the correct format, but contain incorrect info, e.g. two 
stream choices instead of three, or GPA greater than 12.0.
\end{itemize}

Of the two modudles, CalcModule was chosen over ReadAllocationData to test because
CalcModule requires ReadAllocationData to be correct. By testing just CalcModule,
ReadAllocationData is also being tested at the same time because if the first doesn't
work, the second one wouldn't function correctly either.

Also, CalcModule does more interesting things with the data and performs operations on it
making the test cases that would come out of it more unique and less obvious.

\item The problem with using strings for a set of finite elements is that it leaves
more room for error. Although the same functionality could be easily implemented with
strings, using integers or something related would reduces the margin of error in the
program. This new implementation would help catch errors early, rather than miss data
when performing operations later. For example, having 0 = male, and 1 = female, means
that when the data is read from the file, if it is not either a 1 or 0, an error is 
thrown and the mistake could be corrected. If they were instead represented as before,
and the mistake was allowed to propagate, when the average function is called, it would
miss the person with the spelling mistake representing their gender.

\item Dictionaries aren't the only option to implement records and structs in Python,
some other options include tuples, a custom class, collections.namedtuple class, 
typing.NamedTuple class, and the struct.Struct class. Based on the information I learned
from: https://dbader.org/blog/records-structs-and-data-transfer-objects-in-python
I would recomend changing the data structure to either collections.namedtuple or
typing.NamedTuple. The reason for this is because similar to classes, both data
structures allow for using the same format and ensuring that the fields and their 
data are correct. The two structures are also immutable and can be accessed through
unique identifiers. They have all the properties we want of dictionaries but are
safer to use. In terms of implementation in the code, each student would be a
collections.namedtuple, or typing.NamedTuple, and each student would be stored 
in a list.

\item If the list of strings was changed to a tuple, the CalcModule would not to be
changed as the implentation would work the same. If it were changed to another data structure
that is indexed by values, it may not need to be changed, but changing it to one that isn't
would most likely require the code to be altered.

If students were implemented as a custom class or Abstract Data Type (ADT), if the data
structure in the class was changed, the CalcModule code would not need to be altered
because of abstraction. The client, which is the program running, does not need to know
how the data structure is implemented, rather it just calls the method in the class and
uses the value returned by it.

\end{enumerate}

\newpage

\lstset{language=Python, basicstyle=\tiny, breaklines=true, showspaces=false,
  showstringspaces=false, breakatwhitespace=true}
%\lstset{language=C,linewidth=.94\textwidth,xleftmargin=1.1cm}

\def\thesection{\Alph{section}}

\section{Code for ReadAllocationData.py}

\noindent \lstinputlisting{../src/ReadAllocationData.py}

\newpage

\section{Code for CalcModule.py}

\noindent \lstinputlisting{../src/CalcModule.py}

\newpage

\section{Code for testCalc.py}

\noindent \lstinputlisting{../src/testCalc.py}

\newpage

\section{Code for Partner's CalcModule.py}

\noindent \lstinputlisting{../partner/CalcModule.py}

\newpage

\section{Makefile}

\lstset{language=make}
\noindent \lstinputlisting{../Makefile}

\end {document}
